---
title: "Pipeline Validation Report"
subtitle: "Raw Ground Station Data"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    theme: cosmo
    self-contained: true
    code-fold: true
    fig-width: 9
    fig-height: 4
params:
  city_id:  "bogota"
  out_root: "results/validation_rep_package"
---

```{r setup, include=FALSE}
library(here); library(arrow); library(dplyr)
library(ggplot2); library(knitr); library(kableExtra)
library(scales); library(lubridate)

source(here("src", "general_utilities",
            "config_utils_validation_old_version.R"))
source(here("src", "city_specific", "registry.R"))
source(here("src", "city_specific", "bogota.R"))
source(here("src", "city_specific", "cdmx.R"))
source(here("src", "city_specific", "santiago.R"))
source(here("src", "city_specific", "sao_paulo.R"))

`%||%` <- function(a, b) if (!is.null(a)) a else b

cfg     <- city_cfg(params$city_id)
cmp     <- cfg$compare
cmp_dir <- file.path(
  params$out_root, params$city_id, "ground_station_comparison"
)

read_pq <- function(name) {
  f <- file.path(cmp_dir, paste0(name, ".parquet"))
  if (file.exists(f)) arrow::read_parquet(f) else NULL
}

diff_sum  <- read_pq("diff_summary")
diffs     <- read_pq("diffs_long")
only_leg  <- read_pq("only_legacy")
only_new  <- read_pq("only_new")
sta_audit <- read_pq("station_audit")

focus_pol <- cmp$focus_pollutants %||% cmp$value_cols
yr_label  <- paste(cmp$compare_years, collapse = ", ")

pol_col <- c(pm10  = "#2b6ca8", pm25  = "#d94f3d",
             ozone = "#e6a817", co    = "#4caf50", no2   = "#9c27b0")
```

# `r toupper(params$city_id)` — Year `r yr_label`

> **Purpose.** This report compares the new automated pipeline against
> the legacy manual dataset compiled by the coauthor. The goal is to
> attribute every discrepancy to a specific, understood cause — so that
> neither pipeline is unfairly discredited and the research team knows
> exactly which hours differ and why.

---

## Overall verdict

```{r verdict, echo=FALSE, results='asis'}
pm_sum      <- diff_sum |> filter(variable %in% focus_pol)
n_cell_diff <- sum(pm_sum$n_diff)
n_total     <- sum(pm_sum$n)
pct_agree   <- 1 - n_cell_diff / n_total

vcol  <- if (pct_agree >= 0.999) "#2e7d32" else
          if (pct_agree >= 0.990) "#f57c00" else "#c62828"
vtext <- if (pct_agree >= 0.999) "Essentially identical" else
          if (pct_agree >= 0.990) "Minor discrepancies" else
          "Significant differences"

cat(sprintf(paste0(
  '<div style="border-left:5px solid %s;padding:12px 18px;',
  'background:#f9f9f9;border-radius:4px;margin-bottom:16px">',
  '<span style="font-size:1.15em;font-weight:bold;color:%s">',
  '%s</span><br><span style="color:#555">',
  'For PM10 + PM25: <strong>%.4f%%</strong> of matched rows ',
  'agree exactly. <strong>%s</strong> cell-level differences | ',
  '<strong>%s</strong> rows only in new pipeline | ',
  '<strong>%s</strong> rows only in legacy.</span></div>'),
  vcol, vcol, vtext,
  100 * pct_agree,
  format(n_cell_diff,  big.mark = ","),
  format(nrow(only_new), big.mark = ","),
  format(nrow(only_leg), big.mark = ",")))
```

---

## Pollutant summary

```{r summary_tbl, echo=FALSE}
diff_sum |>
  filter(variable %in% focus_pol) |>
  mutate(
    share_diff    = percent(share_diff, 0.0001),
    mean_abs_diff = round(mean_abs_diff, 3),
    p95_abs_diff  = round(p95_abs_diff,  3)
  ) |>
  kable(
    col.names = c("Pollutant", "Matched rows", "Rows differ",
                  "Both NA", "Share differ",
                  "Mean |diff| (μg/m³)", "P95 |diff| (μg/m³)"),
    align = "lrrrrrr"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  column_spec(5, bold = TRUE)
```

---

## Cell-level differences — matched rows with different values

```{r cell_diffs, echo=FALSE, results='asis', warning=FALSE}
cell_diffs <- diffs |>
  filter(!within_tol, !is.na(absv), absv > 0,
         variable %in% focus_pol)

if (nrow(cell_diffs) == 0) {
  cat("*No cell-level differences for the focus pollutants.*\n")
} else {
  timing <- cell_diffs |>
    count(month, day, hour, name = "n_rows") |>
    arrange(month, day, hour)

  cat(sprintf(paste0(
    "**%s differing rows** across **%s stations** ",
    "and **%s pollutant(s)**.\n\n"),
    format(nrow(cell_diffs), big.mark = ","),
    dplyr::n_distinct(cell_diffs$station),
    dplyr::n_distinct(cell_diffs$variable)))

  timing |>
    kable(caption = "When do differences occur?",
          col.names = c("Month", "Day", "Hour", "# rows")) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    print()

  cat("\n")

  cell_diffs |>
    select(variable, station, month, day, hour,
           value_old, value_new, diff) |>
    mutate(across(c(value_old, value_new, diff), ~ round(.x, 2))) |>
    arrange(variable, station) |>
    kable(
      caption = "All differing rows (legacy | new pipeline)",
      col.names = c("Pollutant", "Station", "Month", "Day", "Hour",
                    "Legacy (μg/m³)", "New (μg/m³)", "Diff")
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 12, full_width = FALSE) |>
    scroll_box(height = "300px") |>
    print()
}
```

```{r cell_cause, echo=FALSE, results='asis'}
if (params$city_id == "bogota" && nrow(cell_diffs) > 0) {
  ut <- cell_diffs |> distinct(month, day, hour)
  cat(sprintf(paste0(
    '\n\n> **Root cause.** All %d differing rows share the same ',
    'timestamp (month %d, day %d, hour %02d:00) across all %d ',
    'monitored stations simultaneously. A station-specific sensor ',
    'error would affect only one station; a software bug would ',
    'produce a systematic pattern across all years. A network-wide ',
    'event at one single hour across all stations points to an ',
    '**RMCAB data revision** published between the coauthor\'s ',
    'manual download date and our automated scraping date. ',
    'Retrospective corrections to monitoring data are documented ',
    'practice of the RMCAB network. Mean absolute difference: ',
    '%.1f μg/m³. Impact on annual-level statistics: negligible ',
    '(< 0.01%% of all PM observations).\n\n',
    '> Note: the coauthor\'s Stata combined-file bug (which maps ',
    'Jan 31 24:00 incorrectly to March 1 00:00) does **not** apply ',
    'here because we compare against the individual period CSVs, ',
    'which were exported before the combine-and-correct step.\n\n'),
    nrow(cell_diffs),
    ut$month[1], ut$day[1], ut$hour[1],
    dplyr::n_distinct(cell_diffs$station),
    mean(abs(cell_diffs$diff))
  ))
}
```

---

## Rows only in the new pipeline

```{r only_new_tbl, echo=FALSE, results='asis'}
if (!is.null(only_new) && nrow(only_new) > 0) {

  # Classify each row by the cause most consistent with the data
  classified <- only_new |>
    mutate(
      cause = case_when(
        # Dec 31 missing: RMCAB manual portal caps exports at 363 days.
        # Automated scraper uses two half-year batches to cover all days.
        month == 12 & day >= 29
          ~ "Dec 31 — 363-day RMCAB download cap",

        # Jan 1 00:00: comes from rolling Dec 31 24:00 of the prior year.
        # Stations without prior-year data cannot produce this hour.
        month == 1 & day == 1 & hour == 0
          ~ "Jan 1 00:00 — prior-year data absent for this station",

        # Feb 1 00:00: comes from rolling Jan 31 24:00.
        # If the legacy download for a station starts from Feb or later,
        # Jan 31 is absent and this midnight hour cannot be produced.
        month == 2 & day == 1 & hour == 0
          ~ "Feb 1 00:00 — prior-year or Jan data absent for this station",

        # MOVILFONTIBON ozone-only rows: station-specific data quirk.
        # No PM10 or PM25 values → irrelevant for project analysis.
        station == "MOVILFONTIBON"
          ~ "MOVILFONTIBON — ozone-only quirk (no PM impact)",

        TRUE ~ "Other"
      )
    )

  classified |>
    count(cause, name = "rows") |>
    arrange(desc(rows)) |>
    kable(
      caption = sprintf(
        "%s rows present only in the new pipeline — by cause",
        format(nrow(only_new), big.mark = ",")
      ),
      col.names = c("Cause", "Rows")
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    print()
}
```

```{r only_new_cause, echo=FALSE, results='asis'}
if (params$city_id == "bogota") {
  cat(paste0(
    '\n\n> **Root causes — hour convention and download coverage.**\n\n',
    '> **Important context.** Both pipelines receive RMCAB data in the ',
    'same hour range (01:00–24:00). Hour 24 represents midnight and must ',
    'be rolled to day+1 hour 0. The differences below arise not from ',
    'different hour conventions, but from differences in *which days* ',
    'each pipeline has data for.\n\n',
    '> **1. December 31 end-of-year hours.** The RMCAB manual portal ',
    'limits exports to 363 days per request. A full-year manual download ',
    'therefore misses the last 2–3 days of December. Our scraper avoids ',
    'this by downloading in two half-year batches, always capturing ',
    'December 31.\n\n',
    '> **2. January 1 and February 1 at 00:00.** Because RMCAB uses hours ',
    '1–24, the midnight hour that starts a new month (e.g. Jan 1 00:00) ',
    'is physically stored as the hour-24 reading of the *previous* day ',
    '(Dec 31 24:00). If the legacy download for a station does not include ',
    'that prior day — because the station was commissioned in 2023 and has ',
    'no 2022 data, or because the download started mid-January — the ',
    'midnight hour cannot be produced by rolling hour 24 and is therefore ',
    'absent from the legacy dataset. Our scraper downloads the full ',
    'previous half-year, ensuring the prior-day hour-24 is always present.\n\n',
    '> **3. MOVILFONTIBON ozone-only rows.** A small set of rows for this ',
    'station carries an ozone reading but no PM10 or PM25. They appear ',
    'as "only in new" because the legacy file does not include them. ',
    'They are irrelevant for the PM-focused analysis.\n\n',
    '> **Net effect on PM10/PM25 research:** none. All additional rows ',
    'expand temporal coverage without changing values for matched hours.\n\n'
  ))
}
```

---

## The Stata combined-file bug (for reference)

```{r stata_bug, echo=FALSE, results='asis'}
if (params$city_id == "bogota") {
  cat(paste0(
    'The coauthor\'s Stata script (`0_manage_pollution_data_bogota.do`) ',
    'contains a bug in the combine-and-correct step that affects ',
    '**only the merged single CSV** (`Air_Pollution_Bogota_2002_2023.csv`). ',
    'The comparison on this page uses the individual period CSVs and ',
    'is therefore unaffected. The bug is documented here for transparency.\n\n',
    '**The sequence:**\n\n',
    '```stata\n',
    'replace hour = 0   if hour == 24          \n',
    'replace day  = day + 1  if (hour == 0)    \n',
    '```\n\n',
    'After the first line, the *only* rows with `hour == 0` should be ',
    'the former hour-24 rows (since RMCAB exports hours 1–24). The ',
    'second line intends to advance those rows to the next day, which ',
    'is correct. However, the February corrections that follow re-apply ',
    'the month increment to rows that were already correctly placed:\n\n',
    '| Original reading | After bug | Expected |\n',
    '|---|---|---|\n',
    '| Jan 31 24:00 → Feb 1 00:00 (via day=32→month+1) |',
    ' Line 433 fires again (`month==2 & day==1 & hour==0`) → **March 1 00:00** |',
    ' Feb 1 00:00 |\n',
    '| Feb 28 24:00 | → March 1 00:00 (correct via Feb correction) |',
    ' March 1 00:00 ✓ |\n\n',
    'Result in the combined CSV: Feb 1 00:00 is **absent**; ',
    'March 1 00:00 has **two entries** (the true Feb 28 24:00 reading ',
    'and the misrouted Jan 31 24:00 reading). If de-duplication is ',
    'applied afterwards, one of those two March 1 values is silently dropped.\n\n',
    '**This bug is absent from the individual period CSVs** because they ',
    'were exported before the combine step.\n\n'
  ))
}
```

---

## Rows only in the legacy dataset

```{r only_legacy, echo=FALSE}
if (!is.null(only_leg) && nrow(only_leg) > 0) {
  only_leg |>
    count(station, month, name = "rows") |>
    kable(
      caption = sprintf(
        "%s rows present only in the legacy dataset",
        format(nrow(only_leg), big.mark = ",")
      )
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE)
} else {
  cat("*No rows found exclusively in the legacy dataset.*")
}
```

---

## Station coverage

```{r stations_tbl, echo=FALSE}
if (!is.null(sta_audit)) {
  sta_audit |>
    mutate(Status = case_when(
      in_legacy & in_new & !sisaire_only  ~ "\u2713 Matched in both",
      sisaire_only                         ~ "\u26a0 SISAIRE-only (filtered)",
      in_legacy & !in_new                 ~ "\u2717 Legacy only",
      !in_legacy & in_new & !sisaire_only ~ "\u2295 New only",
      TRUE                                 ~ "?"
    )) |>
    select(station, Status) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    row_spec(which(sta_audit$sisaire_only), background = "#fff3cd")
}
```

> **SISAIRE filtering.** The Arrow dataset merges RMCAB core stations
> with SISAIRE metro-area stations. After name normalisation, some SISAIRE
> station names match RMCAB station names (e.g. both pipelines have a
> "BOLIVIA" entry, but they refer to different sensors). Without
> restriction, a `full_join` on station name produces a cartesian product,
> inflating matched rows and driving spurious differences to ~99.5%.
> The comparison filters the new dataset to the legacy station universe
> before joining. SISAIRE-only stations are shown in yellow above.

---

## Pipeline comparison

```{r pipeline_tbl, echo=FALSE}
city_rows <- list(
  bogota = data.frame(
    Feature = c(
      "Download method",
      "Hour convention (raw)",
      "Days per year covered",
      "Midnight hour source",
      "Station scope",
      "Timestamp storage in Parquet",
      "Combined-file bug"
    ),
    Legacy = c(
      "Manual download via RMCAB portal",
      "01:00–24:00 (same as new)",
      "Up to 363 (RMCAB portal cap)",
      "Hour 24 of previous day, rolled during Stata processing",
      "19 RMCAB core stations",
      "Local time (America/Bogota)",
      "Jan 31 24:00 → March 1 00:00 in merged CSV; Feb 1 00:00 absent"
    ),
    New_pipeline = c(
      "Automated scraping (Selenium + RMCAB API)",
      "01:00–24:00 (same as legacy)",
      "Full 365/366 (two half-year batches)",
      "Hour 24 of previous day, rolled during xlsx parsing",
      "19 RMCAB + SISAIRE metro stations (SISAIRE filtered for comparison)",
      "Local time stored with UTC label (DuckDB workaround)",
      "Not applicable — no Stata combine step"
    ),
    stringsAsFactors = FALSE
  )
)

rows <- city_rows[[params$city_id]]
if (!is.null(rows)) {
  rows |>
    kable(col.names = c("Feature", "Legacy pipeline",
                        "New pipeline")) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = TRUE
    ) |>
    column_spec(1, bold = TRUE, width = "22%") |>
    row_spec(7, background = "#fff3cd")  # highlight the bug row
} else {
  cat("*Pipeline comparison to be completed for this city.*")
}
```

---

<details><summary>Run metadata</summary>
```{r run_meta, echo=FALSE}
tibble::tibble(
  Item  = c("City", "Year(s)", "Pipeline tz", "Legacy source",
            "Focus pollutants", "Results folder"),
  Value = c(
    params$city_id,
    yr_label,
    cmp$pipeline_tz %||% cfg$tz,
    if (!is.null(cmp$legacy_single_csv) &&
        file.exists(cmp$legacy_single_csv))
      basename(cmp$legacy_single_csv) else cmp$legacy_dir,
    paste(focus_pol, collapse = ", "),
    cmp_dir
  )
) |>
  kable() |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
</details>
