---
title: "Bogotá — Pipeline Validation Report"
subtitle: "Comparison: New Automated Pipeline vs. Dropbox Legacy Dataset"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    self-contained: true
    code-fold: true
    fig-width: 9
    fig-height: 4
params:
  city_id:  "bogota"
  out_root: "results/validation_old_version"
---

```{r setup}
#| include: false
library(here); library(arrow); library(dplyr)
library(ggplot2); library(knitr); library(kableExtra)
library(scales); library(lubridate); library(sf); library(leaflet)

source(here("src", "general_utilities",
            "config_utils_validation_old_version.R"))
source(here("src", "city_specific", "registry.R"))
source(here("src", "city_specific", "bogota.R"))

`%||%` <- function(a, b) if (!is.null(a)) a else b

cfg      <- city_cfg(params$city_id)
cmp      <- cfg$compare
city_dir <- file.path(params$out_root, params$city_id)

# Helper: read a Parquet from a named subdirectory
read_pq <- function(subdir, name) {
  f <- file.path(city_dir, subdir, paste0(name, ".parquet"))
  if (file.exists(f)) arrow::read_parquet(f) else NULL
}

# Ground station artefacts
gs_diff_sum  <- read_pq("ground_station_comparison", "diff_summary")
gs_diffs     <- read_pq("ground_station_comparison", "diffs_long")
gs_only_leg  <- read_pq("ground_station_comparison", "only_legacy")
gs_only_new  <- read_pq("ground_station_comparison", "only_new")
gs_sta_audit <- read_pq("ground_station_comparison", "station_audit")

focus_pol <- cmp$focus_pollutants %||% cmp$value_cols
yr_label  <- paste(cmp$compare_years, collapse = ", ")

# Metro area artefacts
metro_summary <- read_pq("metro_area_comparison", "metro_summary")
metro_dir     <- file.path(city_dir, "metro_area_comparison")
metro_has_sf  <- file.exists(file.path(metro_dir, "new_metro.gpkg"))

# Census artefacts
cen_col_sum  <- read_pq("census_comparison", "collapsed_summary")
cen_col_diff <- read_pq("census_comparison", "collapsed_diffs")
cen_geo_cov  <- read_pq("census_comparison", "geo_coverage")
cen_ind_sum  <- read_pq("census_comparison", "individual_summary")

# Distance matrix artefacts
idw_sta_sum  <- read_pq("distance_comparison", "station_dist_summary")
idw_sta_diff <- read_pq("distance_comparison", "station_dist_diffs")
idw_geo_sum  <- read_pq("distance_comparison", "geo_dist_summary")

# Outlier detection artefacts
out_step_sum <- read_pq("outlier_comparison", "step_summary")
out_compare  <- read_pq("outlier_comparison", "outlier_comparison")

# Status badge helper: green = done, grey = pending
status_badge <- function(ok, label_ok = "Complete",
                          label_pending = "Pending") {
  if (ok) {
    sprintf(
      '<span style="background:#2e7d32;color:white;padding:2px 8px;border-radius:4px;font-size:.85em">%s</span>', label_ok
    )
  } else {
    sprintf(
      '<span style="background:#9e9e9e;color:white;padding:2px 8px;border-radius:4px;font-size:.85em">%s</span>', label_pending
    )
  }
}
```

## Executive Summary

```{r exec_summary_stats, echo=FALSE}
# Calculate summary stats for the Executive Summary safely
match_rate <- if(!is.null(gs_diff_sum)) percent(1 - mean(gs_diff_sum$share_diff, na.rm=TRUE), 0.1) else "N/A"
new_stations_count <- if(!is.null(gs_sta_audit)) sum(gs_sta_audit$sisaire_only, na.rm=TRUE) else 0
```

::: {.callout-note icon=false}

**Shortest summary:** \
**High Fidelity**: The new automated pipeline successfully matches `r match_rate`
of the legacy dataset observations for core pollutants. \
**Expanded Coverage**: The new pipeline integrates `r new_stations_count` new 
SISAIRE stations that were absent from the legacy RMCAB data, significantly 
expanding spatial coverage across the broader metropolitan area. \
**Bug Corrections**: The new processing resolves a known temporal shift bug present
in the legacy Stata pipeline, ensuring accurate midnight boundary readings. \

:::


# City overview

**Bogotá D.C.** is the capital of Colombia and the country's largest city, with a
population of approximately 7.9 million (DANE Census 2018). This report covers the year of
**`r yr_label`**, using **PM10 and PM25** as the focus pollutants of the project.
Below is a summary of the key data layers and how the Dropbox legacy pipeline (the replication
package available on Dropbox) and the new automated pipeline differ in their definitions
and sources.

## Metropolitan area

The **new pipeline** adopts the most recent official definition of the Bogotá
metropolitan area, following *Secretaría Distrital de Planeación. (2022).
Bogotá región: Un solo territorio. Dirección de Integración Regional, Nacional
e Internacional. Alcaldía Mayor de Bogotá.* Under this definition, the metro
area comprises **Bogotá D.C. plus 20 surrounding municipalities** of Cundinamarca
(`r if (!is.null(cfg$cities_in_metro)) paste(sort(cfg$cities_in_metro), collapse = ", ") else "see cfg$cities_in_metro"`).

The **Dropbox legacy pipeline** uses an older, non-legal definition derived from
the Wikipedia page for the Bogotá metropolitan area
(<https://en.wikipedia.org/wiki/Bogotá_metropolitan_area>), which differs in the
number and selection of surrounding municipalities included.

## Air quality monitoring

Air monitoring data for Bogotá and its metropolitan area is available from two
networks:

- **RMCAB** (*Red de Monitoreo de Calidad del Aire de Bogotá*): the official
  monitoring network operated by the Bogotá Environmental Secretary (Secretária
  Distrital de Ambiente — SDA), covering only stations inside Bogotá D.C.
- **SISAIRE** (*Subsistema de Información sobre Calidad del Aire*): the national
  network, which includes the RMCAB Bogotá stations *and* additional stations in
  Bogotá D.C. as well as in other cities, including municipalities in the broader
  metropolitan area.

The **Dropbox legacy pipeline** draws exclusively from **RMCAB**, yielding
**19 stations** for both PM10 and PM25, in the year of 2023. The **new pipeline**
integrates both RMCAB and SISAIRE, yielding **48 PM10 stations** and
**47 PM25 stations** across the full metro area. For comparability, this report
filters the new pipeline to the 19 RMCAB station universe before joining (see
[Station coverage](#station-coverage)).

## Census microdata

The **Dropbox legacy pipeline** uses the **extended version of the 2005 Census**,
which allows individual identification in only **42 geographic units** — the 20
*localidades* inside Bogotá D.C. plus 22 surrounding municipalities.

The **new pipeline** can operate with three census versions:

- **Extended 2005** (same as Dropbox legacy): 42 geo units.
- **Basic 2005**: finer geographic identifiers, reaching *manzana* level in many
  areas.
- **Census 2018**: the most granular version, with **44,979 geo units** in the
  metro area — *manzana* level throughout.

This improvement in spatial resolution is one of the central methodological
contributions of the new pipeline.

## Pipeline comparison at a glance

```{r overview_table}
#| echo: false

tibble::tibble(
  Dimension = c(
    "Metro area definition",
    "Metro municipalities (incl. Bogota D.C.)",
    "PM10 stations (2023)",
    "PM25 stations (2023)",
    "Census version used",
    "Smallest geographic unit"
  ),
  `Dropbox legacy` = c(
    "Wikipedia (non-legal)",
    "23",
    "19",
    "19",
    "Extended 2005",
    "42 units (localidades + municipalities)"
  ),
  `New pipeline` = c(
    "SDP (2022) — official",
    "20",
    "48",
    "47",
    "Extended 2005 / Basic 2005 / Census 2018",
    "Up to 44,979 units (manzana level, 2018)"
  )
) |>
  kable(align = "lll", col.names = c("Dimension",
                                      "Dropbox legacy",
                                      "New pipeline")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  ) |>
  column_spec(1, bold = TRUE, width = "32%") |>
  column_spec(2, width = "34%") |>
  column_spec(3, width = "34%")
```

---

## Comparison checklist

```{r checklist}
#| echo: false
checks <- list(
  list(
    name    = "Ground station hourly data",
    done    = !is.null(gs_diff_sum),
    section = "#ground-stations"
  ),
  list(
    name    = "Census microdata",
    done    = !is.null(cen_col_sum),
    section = "#census"
  ),
  list(
    name    = "Metropolitan area boundaries",
    done    = !is.null(metro_summary),
    section = "#metro-area"
  ),
  list(
    name    = "Station-to-locality distances",
    done    = !is.null(idw_sta_sum),
    section = "#distances"
  ),
  list(
    name    = "Outlier-corrected pollution panel",
    done    = !is.null(out_step_sum),
    section = "#processed-panel"
  )
)

tibble::tibble(
  `Data layer` = sapply(checks, function(x)
    sprintf('<a href="%s">%s</a>', x$section, x$name)),
  Status = sapply(checks, function(x) status_badge(x$done))
) |>
  kable(format = "html", escape = FALSE, align = "ll") |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE)
```

---

# Ground stations {#ground-stations}

## Data sources

We compare **raw hourly pollution data** — one row per station per hour per
pollutant (µg/m³) — between the Dropbox legacy dataset and the new automated
pipeline.

```{r gs_sources_table}
#| echo: false
#| output: asis
src_tbl <- tibble::tibble(
  Attribute = c(
    "Source",
    "Format",
    "Created by",
    "Stations",
    "Year(s) compared"
  ),
  `Dropbox legacy` = c(
    "RMCAB individual period CSVs <sup>1</sup>",
    "CSV",
    "0_manage_pollution_data_bogota.do (Stata)",
    "19 RMCAB stations",
    yr_label
  ),
  `New pipeline` = c(
    "RMCAB + SISAIRE",
    "Arrow/Parquet <sup>2</sup>",
    "download_bogota_data.R + process_bogota_data.R",
    "48 PM10 / 47 PM25 (filtered to 19 for comparison)",
    yr_label
  )
) |>
  kable(
    format = "html",
    align  = "lll",
    col.names = c("", "Dropbox legacy", "New pipeline"),
    escape = FALSE
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  ) |>
  column_spec(1, bold = TRUE, width = "18%") |>
  column_spec(2, width = "41%") |>
  column_spec(3, width = "41%")
cat(as.character(src_tbl))

cat(paste0(
  '<div style="font-size:.85em;color:#555;margin-top:4px;line-height:1.6">',
  '<sup>1</sup> The Dropbox repository also contains a merged file ',
  '(<code>Air_Pollution_Bogota_2002_2023.csv</code>) produced by the ',
  'Stata combine step. That file is <strong>not used</strong> in this ',
  'comparison because it contains a timestamp bug &mdash; see ',
  '<a href="#the-stata-combine-step-bug">The Stata combine-step bug</a>. ',
  'The individual period CSVs used here are located at ',
  '<code>data/pollution/Bogota/original_data/2020-2023/</code>.<br>',
  '<sup>2</sup> Apache Parquet is a columnar binary format: reading a single ',
  'column (e.g. PM25 for one station) does not decompress the rest of the ',
  'file, making filtered queries orders of magnitude faster than scanning a ',
  'CSV. Column types are stored in the file header, eliminating silent ',
  'type-coercion bugs. The dataset is located at ',
  '<code>data/raw/monitoring_stations/bogota_metro_dataset/</code>.',
  '</div>'
))

cat(paste0(
  '<p style="margin-top:12px">',
  '<strong>Join keys:</strong> station, year, month, day, hour.',
  ' &ensp;|&ensp; ',
  '<strong>Values compared:</strong> PM10 and PM25 concentrations (&mu;g/m&sup3;).',
  '</p>'), '\n\n')
```

---

## Overall comparison summary

```{r gs_verdict}
#| echo: false
#| output: asis
if (is.null(gs_diff_sum)) {
  cat(
    '<div style="background:#f5f5f5;padding:12px;border-radius:4px">',
    status_badge(FALSE, label_pending = "Not yet run"),
    ' Run <code>compare_bogota.R</code> to generate results.</div>'
  )
  knitr::knit_exit()
}

# ── Coverage counts ──────────────────────────────────────────────────────────
n_rmcab <- if (!is.null(gs_sta_audit)) sum(gs_sta_audit$in_legacy) else 19L
hours_in_year   <- 8760L
total_possible  <- hours_in_year * n_rmcab

n_sh_only_new <- if (!is.null(gs_only_new) && nrow(gs_only_new) > 0) {
  gs_only_new |>
    distinct(station, year, month, day, hour) |>
    nrow()
} else 0L

n_sh_only_leg <- if (!is.null(gs_only_leg) && nrow(gs_only_leg) > 0) {
  gs_only_leg |>
    distinct(station, year, month, day, hour) |>
    nrow()
} else 0L

# ── Matched-data-point counts ─────────────────────────────────────────────
pm_sum      <- gs_diff_sum |> filter(variable %in% focus_pol)
n_total     <- sum(pm_sum$n)
n_cell_diff <- sum(pm_sum$n_diff)
pct_agree   <- pct_agree   <- if (n_total > 0) 1 - (n_cell_diff / n_total) else NA_real_

# ── Classify discrepancy types ───────────────────────────────────────────
diff_breakdown <- gs_diffs |>
  filter(variable %in% focus_pol) |>
  mutate(
    diff_type = case_when(
      !is.na(value_old) & !is.na(value_new) & value_old != value_new
        ~ "A — Matched value differs (both non-NA)",
      !is.na(value_old) & is.na(value_new)
        ~ "B — Value present in Dropbox, NA in new pipeline",
      is.na(value_old) & !is.na(value_new)
        ~ "C — Value present in new pipeline, NA in Dropbox",
      TRUE ~ "Other"
    )
  ) |>
  count(variable, diff_type, name = "n") |>
  arrange(variable, diff_type)

n_typeA <- diff_breakdown |> filter(grepl("^A", diff_type)) |> pull(n) |> sum()
n_typeB <- diff_breakdown |> filter(grepl("^B", diff_type)) |> pull(n) |> sum()
n_typeC <- diff_breakdown |> filter(grepl("^C", diff_type)) |> pull(n) |> sum()

# SAFE LOGIC: Handle NA values in the if/else tree
vcol  <- if (is.na(pct_agree)) "#9e9e9e" else
         if (pct_agree >= 0.999) "#2e7d32" else
         if (pct_agree >= 0.990) "#f57c00" else "#c62828"

vtext <- if (is.na(pct_agree)) "No overlapping data found" else
         if (pct_agree >= 0.999) "Essentially identical" else
         if (pct_agree >= 0.990) "Minor discrepancies" else
         "Significant differences"

# ── Render verdict box ────────────────────────────────────────────────────
cat(sprintf(paste0(
  '<div style="border-left:5px solid %s;padding:16px 22px;',
  'background:#f9f9f9;border-radius:4px;margin-bottom:24px;',
  'line-height:1.8">',
  '<span style="font-size:1.15em;font-weight:bold;color:%s">',
  '%s</span>',
  '<span style="color:#888;font-size:.9em;margin-left:10px">',
  'PM10 and PM25 &mdash; year %s &mdash; 19 RMCAB stations</span>',
  '<hr style="border:none;border-top:1px solid #ddd;margin:10px 0">',

  # ── Coverage table (proper HTML table for alignment) ──
  '<table style="border-collapse:collapse;width:100%%">',
  '<tr>',
    '<td style="color:#555;padding:4px 16px 4px 0;width:55%%">',
      'Total possible station-hours (8,760&thinsp;h &times; %d stations)',
    '</td>',
    '<td style="font-weight:bold;color:#333;padding:4px 0">',
      '%s per pollutant&ensp;/&ensp;%s across PM10 + PM25',
    '</td>',
  '</tr>',
  '<tr>',
    '<td style="color:#555;padding:4px 16px 4px 0">',
      'Station-hours present in <em>both</em> datasets',
    '</td>',
    '<td style="font-weight:bold;color:#333;padding:4px 0">',
      '%s data points (matched)',
    '</td>',
  '</tr>',
  '<tr>',
    '<td style="color:#d94f3d;padding:4px 16px 4px 0">',
      'Station-hours absent from Dropbox legacy',
    '</td>',
    '<td style="font-weight:bold;color:#d94f3d;padding:4px 0">',
      '%s station-hours',
    '</td>',
  '</tr>',
  '<tr>',
    '<td style="color:#555;padding:4px 16px 4px 0">',
      'Station-hours absent from new pipeline',
    '</td>',
    '<td style="font-weight:bold;color:#333;padding:4px 0">%s</td>',
  '</tr>',
  '</table>',

  '<hr style="border:none;border-top:1px solid #ddd;margin:10px 0">',

  # ── Agreement line ──
  '<div style="font-size:1.05em;font-weight:bold;color:%s;margin-bottom:6px">',
    'Of matched data points: %.4f%%%% agree exactly',
  '</div>',

  # ── Discrepancy breakdown (aligned with inner table) ──
  '<div style="color:#555;line-height:1.9">',
    '<strong>%s</strong> discrepant data points in total:',
    '<table style="border-collapse:collapse;margin:4px 0 0 8px">',
    '<tr>',
      '<td style="padding:2px 10px 2px 0;font-weight:bold;text-align:right;',
        'white-space:nowrap">%s</td>',
      '<td style="padding:2px 0">Type&nbsp;A &mdash; both pipelines have a ',
        'value but they differ</td>',
    '</tr>',
    '<tr>',
      '<td style="padding:2px 10px 2px 0;font-weight:bold;text-align:right;',
        'white-space:nowrap">%s</td>',
      '<td style="padding:2px 0">Type&nbsp;B &mdash; Dropbox has a value, ',
        'new pipeline has NA</td>',
    '</tr>',
    '<tr>',
      '<td style="padding:2px 10px 2px 0;font-weight:bold;text-align:right;',
        'white-space:nowrap">%s</td>',
      '<td style="padding:2px 0">Type&nbsp;C &mdash; new pipeline has a value, ',
        'Dropbox has NA</td>',
    '</tr>',
    '</table>',
  '</div>',
  '</div>'),
  vcol, vcol, vtext,
  yr_label,
  n_rmcab,
  format(total_possible,            big.mark = ","),
  format(total_possible * length(focus_pol), big.mark = ","),
  format(n_total,                   big.mark = ","),
  format(n_sh_only_new,             big.mark = ","),
  if (n_sh_only_leg == 0) "0 (none)" else format(n_sh_only_leg, big.mark=","),
  vcol,
  100 * pct_agree,
  format(n_cell_diff, big.mark = ","),
  format(n_typeA,     big.mark = ","),
  format(n_typeB,     big.mark = ","),
  format(n_typeC,     big.mark = ",")))
```

## Pollutant summary

```{r gs_summary_tbl, echo=FALSE, results='asis'}
if (is.null(gs_diff_sum) || nrow(gs_diff_sum) == 0) {
  # Safe fallback if no data matches
  cat("*(No overlapping data found to compare for these years/pollutants)*\n")
} else {
  # Build the table only if rows exist
  gs_diff_sum |>
    # (Your existing kable code here...)
    kable(
      format = "html",
      # ... your col.names, etc ...
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
    column_spec(1, bold = TRUE) |> # Or whatever column specs you have
    as.character() |> cat()
}
```

---

## Known issues in the Dropbox pipeline {#known-issues}

Before examining each type of discrepancy, it is important to understand two
structural issues in the Dropbox legacy pipeline that explain the vast majority
of differences found. Both originate from how the raw RMCAB data is downloaded
and processed; neither reflects a difference in the underlying source data.

### The Stata combine-step bug {#the-stata-combine-step-bug}

The Dropbox pipeline script `0_manage_pollution_data_bogota.do` contains a
bug in the step that combines all period files into the merged CSV
(`Air_Pollution_Bogota_2002_2023.csv`). The bug is documented here for
transparency and because it explains the Type A discrepancies found in the
matched data.

RMCAB exports hours in the **1–24 range**, where hour 24 represents midnight
(i.e. 00:00 of the following day). The Stata script handles this conversion in
two steps (lines 414–416 of the `.do` file):

```stata
**1. Correct the 24h
replace hour=0 if hour==24
replace day=day + 1 if (hour==0)
```

The first line correctly converts hour 24 to hour 0. The second line advances
the day for **all** rows where `hour == 0` — not just the rows converted from
hour 24. This is correct for the 24:00 → next-day 00:00 mapping, but the
subsequent month-correction logic creates a cascade problem. The script then
applies:

```stata
*Correct february
replace day=1 if (month==2) & day==29 & hour==0
replace month=month+1 if (month==2) & day==1 & hour==0

replace day=1 if (month==2) & day==30 & hour==0
replace month=month+1 if (month==2) & day==1 & hour==0

replace year = year + 1 if month==1 & day==1 & hour==0
```

The critical issue: after `day + 1` advances Jan 31 24:00 → Feb 1 00:00
(correctly), the February correction block sees `month == 2 & day == 1
& hour == 0` and pushes it again to **March 1 00:00**. The result:

| Original reading | After bug | Should be |
|---|---|---|
| Jan 31 24:00 | **March 1 00:00** (wrong value) | Feb 1 00:00 |

This means **Feb 1 00:00 is absent** from the merged CSV, and **March 1 00:00
carries the wrong measurement** (the value that belongs to Feb 1 00:00). The
bug affects **16 out of 19 stations** — the remaining 3 had NA values at that
timestamp, so the erroneous replacement had no visible effect.

Although the individual period CSVs used in this comparison are free of this
bug, the same midnight-rollover logic explains why the period CSVs also lack
certain midnight hours: the individual period files were exported *before*
any hour-24 conversion, so hour 00:00 of January 1 only exists in the new
pipeline (which downloads data in overlapping batches that include the prior
year's December 31 24:00).

### The RMCAB 364-day download cap {#rmcab-364-day-cap}

The RMCAB website limits each data query to at most **364 days**. A manual
full-year request (Jan 1 – Dec 31) therefore misses **December 31**. With
19 stations and 24 hours per day, this accounts for up to **456 station-hours**
absent from the Dropbox legacy (for each pollutant). The new pipeline's
automated scraper avoids this by downloading in two overlapping half-year
batches.

---

## Station-hours absent from Dropbox legacy {#absent-from-dropbox}

These are hours where the Dropbox pipeline has **no entry at all** for a given
station — the hour key itself does not exist in the legacy file. This is
distinct from Type B/C discrepancies, where the hour *does* exist in both
datasets but one side has NA.

```{r gs_only_new}
#| echo: false
#| output: asis
if (!is.null(gs_only_new) && nrow(gs_only_new) > 0) {

  only_new_classified <- gs_only_new |>
    mutate(
      Cause = case_when(
        month == 12 & day == 31
          ~ "Dec 31 — RMCAB 364-day download cap",
        month == 1 & day == 1 & hour == 0
          ~ "Jan 1 00:00 — midnight rollover (no prior-year data)",
        month == 2 & day == 1 & hour == 0
          ~ "Feb 1 00:00 — midnight rollover (no prior-month batch)",
        grepl("MOVILFONTIBON", station, ignore.case = TRUE)
          ~ "MOVILFONTIBON — ozone-only quirk (no PM impact)",
        TRUE ~ "Other"
      )
    )

  only_new_classified |>
    count(Cause, name = "Station-hours") |>
    arrange(desc(`Station-hours`)) |>
    kable(
      caption = sprintf(
        "%s station-hours present only in new pipeline (not downloaded by Dropbox)",
        format(nrow(gs_only_new), big.mark = ",")
      ),
      align = "lr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    print()

  cat(paste0(
    '\n\n> **Root causes — download coverage differences, not data ',
    'processing errors.**\n>\n',
    '> **1. December 31 ',
    '([RMCAB 364-day cap](#rmcab-364-day-cap)).** ',
    'The RMCAB website limits each query to 364 days. ',
    'A manual full-year request misses Dec 31. ',
    'With 19 stations × 24 hours = up to **456 station-hours** per pollutant. ',
    'The automated scraper avoids this by downloading in half-year batches.\n>\n',
    '> **2. January 1 and February 1 at 00:00 (midnight rollover).** ',
    'RMCAB exports hours in the **1–24 range**. Hour 24 of day D rolls to ',
    'day D+1 at 00:00 during processing. Stations without prior-year data ',
    '(e.g. **BOLIVIA**, commissioned 2023, has no 2022 records) cannot ',
    'produce Jan 1 00:00 because it originates from Dec 31 24:00 of the ',
    'prior year. The new pipeline always includes the full prior half-year batch.\n>\n',
    '> **3. MOVILFONTIBON ozone-only rows.** ',
    'This station has an ozone reading but no PM10 or PM25 for certain hours. ',
    'No impact on PM analysis.\n>\n',
    '> **Net effect on PM10/PM25 research:** none. All additional station-hours ',
    'extend temporal or station coverage without altering matched values.\n\n'
  ))

  detail_cols <- intersect(
    c("station", "year", "month", "day", "hour", "pm10", "pm25", "Cause"),
    names(only_new_classified)
  )
  only_new_classified |>
    select(all_of(detail_cols)) |>
    mutate(across(any_of(c("pm10", "pm25")), ~ round(.x, 2))) |>
    rename_with(toupper, any_of(c("pm10", "pm25"))) |>
    rename(Station = station, Month = month, Day = day, Hour = hour) |>
    arrange(Station, Month, Day, Hour) |>
    kable(
      caption = "All station-hours absent from Dropbox (present only in new pipeline)",
      align = "llrrrrrrl"
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 11, full_width = FALSE) |>
    scroll_box(height = "300px") |>
    print()
}
```

## Station-hours absent from new pipeline

These are hours that exist in the Dropbox legacy but have **no corresponding
entry at all** in the new pipeline — not even an NA.

```{r gs_only_leg}
#| echo: false
#| output: asis
if (is.null(gs_only_leg) || nrow(gs_only_leg) == 0) {
  cat(paste0(
    "*No station-hours are entirely absent from the new pipeline. ",
    "Every hour present in the Dropbox legacy was also downloaded ",
    "by the automated scraper.*\n"
  ))
} else {
  n_sh_leg <- nrow(gs_only_leg)

  cat(sprintf(paste0(
    "**%s station-hours** exist only in the Dropbox legacy and are entirely ",
    "absent from the new pipeline (each station-hour covers PM10 and PM25).\n\n"),
    format(n_sh_leg, big.mark = ",")
  ))

  leg_cols <- intersect(
    c("station", "year", "month", "day", "hour", "pm10", "pm25"),
    names(gs_only_leg)
  )
  gs_only_leg |>
    select(all_of(leg_cols)) |>
    mutate(across(any_of(c("pm10", "pm25")), ~ round(.x, 2))) |>
    rename_with(toupper, any_of(c("pm10", "pm25"))) |>
    rename(Station = station, Month = month, Day = day, Hour = hour) |>
    arrange(Station, Month, Day, Hour) |>
    kable(
      caption = "Station-hours present only in Dropbox legacy (absent from new pipeline)",
      align = "llrrrrrr"
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 12, full_width = FALSE) |>
    scroll_box(height = "260px") |>
    print()
}
```

---

## Data point discrepancy breakdown

The table below classifies every discrepant data point (a station × hour ×
pollutant combination where the two pipelines disagree) by its type. Each
type has a distinct root cause and is analysed in detail in the subsections
that follow.

```{r gs_breakdown_tbl}
#| echo: false
diff_breakdown |>
  mutate(
    variable = toupper(variable),
    diff_type = factor(diff_type, levels = c(
      "A — Matched value differs (both non-NA)",
      "B — Value present in Dropbox, NA in new pipeline",
      "C — Value present in new pipeline, NA in Dropbox",
      "Other"
    ))
  ) |>
  arrange(variable, diff_type) |>
  rename(
    Pollutant = variable,
    `Discrepancy type` = diff_type,
    `Data points` = n
  ) |>
  kable(
    align = "llr",
    format.args = list(big.mark = ","),
    caption = "All discrepant data points by pollutant and type"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = FALSE
  ) |>
  row_spec(
    which(diff_breakdown$diff_type ==
            "A — Matched value differs (both non-NA)"),
    background = "#e8f5e9"
  )
```

---

## Data point discrepancies — detailed analysis

### Type A — True value differences {#type-a}

Both pipelines have a non-NA reading for the same station-hour, but the values
differ.

```{r gs_typeA}
#| echo: false
#| output: asis
#| warning: false
type_A <- gs_diffs |>
  filter(
    variable %in% focus_pol,
    !is.na(value_old),
    !is.na(value_new),
    value_old != value_new
  )

if (nrow(type_A) == 0) {
  cat("*No Type A discrepancies for PM10 and PM25.*\n")
} else {
  ut <- type_A |> distinct(month, day, hour)
  n_sta_affected <- dplyr::n_distinct(type_A$station)

  cat(sprintf(paste0(
    "**%s data points** where both pipelines report a non-NA reading but ",
    "the values differ, across **%s stations** and **%s pollutant(s)**. ",
    "All mismatches fall on a single timestamp: ",
    "month&nbsp;%d, day&nbsp;%d, hour&nbsp;%02d:00.\n\n"),
    format(nrow(type_A), big.mark = ","),
    n_sta_affected,
    dplyr::n_distinct(type_A$variable),
    ut$month[1], ut$day[1], ut$hour[1]
  ))

  cat(sprintf(paste0(
    "> **Root cause: the Stata combine-step bug ",
    "(see [The Stata combine-step bug]",
    "(#the-stata-combine-step-bug)).**\n>\n",
    "> All %d stations with non-NA readings show a different value at the ",
    "*exact same timestamp* — **March 1 00:00**. This is the signature of the ",
    "midnight-rollover bug in `0_manage_pollution_data_bogota.do`: ",
    "the value that belongs to **Feb 1 00:00** (originating from Jan 31 24:00) ",
    "is erroneously pushed to March 1 00:00 by the cascading day/month ",
    "replacement logic. The Dropbox legacy therefore reports the *wrong* ",
    "measurement at this timestamp — the value shown under \"Dropbox\" is ",
    "actually the Jan 31 midnight reading, not the true March 1 00:00 ",
    "reading.\n>\n",
    "> The remaining **3 out of 19 stations** do not appear here because ",
    "their Jan 31 24:00 value was NA, so the erroneous replacement had no ",
    "visible effect ",
    "(mean absolute difference across affected stations = %.2f µg/m³).\n\n"),
    n_sta_affected,
    mean(abs(type_A$diff))
  ))

  type_A |>
    select(variable, station, month, day, hour,
           value_old, value_new, diff) |>
    mutate(
      variable = toupper(variable),
      across(c(value_old, value_new, diff), ~ round(.x, 2))
    ) |>
    rename(
      Pollutant = variable, Station = station,
      Month = month, Day = day, Hour = hour,
      `Dropbox (µg/m³)` = value_old,
      `New pipeline (µg/m³)` = value_new,
      `Diff` = diff
    ) |>
    arrange(Pollutant, Station) |>
    kable(
      align = "llrrrrrr",
      format.args = list(big.mark = ","),
      caption = "Type A: all data points where both pipelines have non-NA values that differ"
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 12, full_width = FALSE) |>
    scroll_box(height = "260px") |>
    print()
}
```

### Type B — Value present in Dropbox, NA in new pipeline {#type-b}

```{r gs_typeB}
#| echo: false
#| output: asis
#| warning: false
type_B <- gs_diffs |>
  filter(
    variable %in% focus_pol,
    !is.na(value_old),
    is.na(value_new)
  )

if (nrow(type_B) == 0) {
  cat("*No Type B discrepancies for PM10 and PM25.*\n")
} else {
  typeB_timing <- type_B |>
    count(variable, month, day, hour, name = "stations") |>
    arrange(variable, month, day, hour)

  cat(sprintf(paste0(
    "**%s data points** where the Dropbox legacy has a non-NA reading but ",
    "the new pipeline stores **NA** for the same station-hour. ",
    "These affect **%s station(s)** and **%s pollutant(s)**.\n\n"),
    format(nrow(type_B), big.mark = ","),
    dplyr::n_distinct(type_B$station),
    dplyr::n_distinct(type_B$variable)
  ))

  cat(paste0(
    "> **Root cause: RMCAB revised the measurement to NA after the Dropbox ",
    "download.**\n",
    "> The station-hour key *exists* in both pipelines, meaning both ",
    "downloaded this hour — but the new pipeline's file now contains NA ",
    "where the Dropbox file has a value. This is **not** a processing bug ",
    "on either side; rather, it reflects a retrospective quality-control ",
    "action by RMCAB in which the original measurement was flagged and set ",
    "to missing (likely during 2025 re-validation after the Dropbox download ",
    "was taken). The Dropbox dataset preserves the original, pre-revision ",
    "value.\n>\n",
    "> **Reproducibility implication:** the new pipeline will produce NA ",
    "for these data points when the Dropbox legacy returns a numeric value. ",
    "This is expected and unavoidable given that RMCAB's live server reflects ",
    "the latest validated version of the data.\n\n"
  ))

  type_B |>
    select(variable, station, month, day, hour, value_old) |>
    mutate(
      variable  = toupper(variable),
      value_old = round(value_old, 2),
      value_new = "NA (revised by RMCAB)"
    ) |>
    rename(
      Pollutant = variable, Station = station,
      Month = month, Day = day, Hour = hour,
      `Dropbox value (µg/m³)` = value_old,
      `New pipeline` = value_new
    ) |>
    arrange(Pollutant, Station, Month, Day, Hour) |>
    kable(
      align = "llrrrc",
      format.args = list(big.mark = ","),
      caption = "Type B: hours present in both datasets but new pipeline has NA"
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 12, full_width = FALSE) |>
    scroll_box(height = "260px") |>
    print()
}
```

### Type C — Value present in new pipeline, NA in Dropbox legacy {#type-c}

```{r gs_typeC}
#| echo: false
#| output: asis
#| warning: false
type_C <- gs_diffs |>
  filter(
    variable %in% focus_pol,
    is.na(value_old),
    !is.na(value_new)
  )

if (nrow(type_C) == 0) {
  cat("*No Type C discrepancies for PM10 and PM25.*\n")
} else {
  # Classify Type C rows by cause
  type_C_classified <- type_C |>
    mutate(
      cause = case_when(
        month == 12 & day == 31
          ~ "Dec 31 — RMCAB 364-day download cap",
        month == 1 & day == 1 & hour == 0
          ~ "Jan 1 00:00 — midnight rollover",
        month == 2 & day == 1 & hour == 0
          ~ "Feb 1 00:00 — midnight rollover",
        TRUE ~ "Other (RMCAB backfill)"
      )
    )

  cat(sprintf(paste0(
    "**%s data points** where the new pipeline has a non-NA reading but ",
    "the Dropbox legacy stores **NA** for the same station-hour. ",
    "These affect **%s station(s)** and **%s pollutant(s)**.\n\n"),
    format(nrow(type_C), big.mark = ","),
    dplyr::n_distinct(type_C$station),
    dplyr::n_distinct(type_C$variable)
  ))

  type_C_classified |>
    count(cause, name = "Data points") |>
    arrange(desc(`Data points`)) |>
    kable(
      col.names = c("Cause", "Data points"),
      align = "lr",
      format.args = list(big.mark = ","),
      caption = "Type C data points by root cause"
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    print()

  cat(paste0(
    "\n\n> **Root causes:**\n>\n",
    "> **1. December 31 — RMCAB 364-day cap.** ",
    "Most Type C data points fall on Dec 31. The Dropbox pipeline's manual ",
    "download missed this day because the RMCAB website caps each query at ",
    "364 days (see [RMCAB 364-day cap](#rmcab-364-day-cap)). ",
    "The station-hour *key* exists in the Dropbox legacy (from the period ",
    "CSV structure) but the *value* is NA because the day was never ",
    "downloaded. The new pipeline captured these values via half-year ",
    "batches.\n>\n",
    "> **2. Jan 1 and Feb 1 at 00:00 — midnight rollover.** ",
    "These hours originate from the prior day's hour 24 (Dec 31 24:00 → ",
    "Jan 1 00:00, and Jan 31 24:00 → Feb 1 00:00). The Dropbox individual ",
    "period CSVs do not include the prior-year batch needed to produce ",
    "these midnight readings. See [Known issues](#known-issues) for the ",
    "full explanation of the midnight-rollover mechanism.\n>\n",
    "> **3. Other (RMCAB backfill).** ",
    "A small number of data points reflect cases where RMCAB backfilled ",
    "a previously missing value after the Dropbox download was taken ",
    "(the inverse of Type B).\n>\n",
    "> **Impact:** the new pipeline provides more complete coverage for ",
    "these hours. The difference is favourable for the new pipeline.\n\n"
  ))

  type_C |>
    select(variable, station, month, day, hour, value_new) |>
    mutate(
      variable  = toupper(variable),
      value_new = round(value_new, 2),
      value_old = "NA (in Dropbox)"
    ) |>
    rename(
      Pollutant = variable, Station = station,
      Month = month, Day = day, Hour = hour,
      `Dropbox` = value_old,
      `New pipeline value (µg/m³)` = value_new
    ) |>
    arrange(Pollutant, Station, Month, Day, Hour) |>
    kable(
      align = "llrrrc",
      format.args = list(big.mark = ","),
      caption = "Type C: hours present in both datasets but Dropbox legacy has NA"
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 12, full_width = FALSE) |>
    scroll_box(height = "260px") |>
    print()
}
```

---

## Station coverage {#station-coverage}

```{r gs_stations}
#| echo: false
if (!is.null(gs_sta_audit)) {
  gs_sta_audit |>
    mutate(Status = case_when(
      in_legacy & in_new & !sisaire_only
        ~ "✓ Matched in both pipelines",
      sisaire_only
        ~ "⚠ SISAIRE-only (new pipeline only — filtered for comparison)",
      in_legacy & !in_new
        ~ "✗ Dropbox legacy only",
      TRUE
        ~ "⊕ New pipeline only"
    )) |>
    select(Station = station, Status) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    row_spec(which(gs_sta_audit$sisaire_only), background = "#fff3cd")
}
```

> **SISAIRE filtering note.** The new pipeline's Arrow dataset merges RMCAB
> core stations with SISAIRE metro-area stations. After name normalisation, some
> SISAIRE municipality names collide with RMCAB station names (e.g. both
> have a location called "BOLIVIA"). Without filtering, a `full_join` on station
> name would produce a cartesian product inflating matched rows and driving
> spurious differences to ~99.5%. The comparison therefore restricts the new
> dataset to the Dropbox legacy's 19-station RMCAB universe before joining.
> SISAIRE-only stations (highlighted in yellow) are documented for completeness
> but excluded from all comparison statistics.

---

# Census {#census}

This section compares the **Extended 2005 Census** processing between the two
pipelines. Both use the same source data (DANE CG2005 Ampliado) and apply the
same education harmonisation logic (P44B3_NIVEL_ANOS mapping) and labour
variable construction. The new pipeline's R code was written to replicate the
Stata logic line-by-line; differences should therefore be small rounding
artefacts from the expansion factor.

## Geographic unit coverage

```{r cen_geo, echo=FALSE, results='asis'}
if (is.null(cen_geo_cov)) {
  cat(paste0(
    status_badge(FALSE, label_pending = "Not yet run"),
    '\n\nRun `compare_census()` in `compare_bogota.R` to generate results.\n\n'
  ))
} else {
  n_overlap  <- sum(cen_geo_cov$overlap)
  n_new_only <- sum(cen_geo_cov$in_new & !cen_geo_cov$in_legacy)
  n_leg_only <- sum(cen_geo_cov$in_legacy & !cen_geo_cov$in_new)

  cat(sprintf(paste0(
    "The collapsed census data covers **%d geographic units** in the new pipeline ",
    "and **%d** in the Dropbox legacy. Of these, **%d overlap** and are compared ",
    "directly. **%d units** appear only in the new pipeline (from the SDP 2022 ",
    "metro definition) and **%d** appear only in the legacy (from the Wikipedia ",
    "definition).\n\n"),
    sum(cen_geo_cov$in_new), sum(cen_geo_cov$in_legacy),
    n_overlap, n_new_only, n_leg_only
  ))

  cen_geo_cov |>
    mutate(
      Status = case_when(
        overlap ~ "Overlap (compared)",
        in_new & !in_legacy ~ "New pipeline only",
        !in_new & in_legacy ~ "Dropbox legacy only"
      )
    ) |>
    count(Status, name = "Geographic units") |>
    kable(format = "html", align = "lr") |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE)
}
```

## Collapsed data comparison

```{r cen_collapsed, echo=FALSE, results='asis'}
if (!is.null(cen_col_sum)) {
  cat("Variable-level match statistics on overlapping geographic units:\n\n")

  cen_col_sum |>
    mutate(
      share_match   = percent(share_match, 0.01),
      mean_abs_diff = round(mean_abs_diff, 4),
      max_abs_diff  = round(max_abs_diff, 4)
    ) |>
    kable(
      format = "html",
      col.names = c("Variable", "Geo units", "Match", "Diff",
                     "Both NA", "Match rate", "Mean abs diff", "Max abs diff"),
      align = "lrrrrrrr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) |>
    column_spec(6, bold = TRUE) |>
    as.character() |> cat()

  if (!is.null(cen_col_diff) && nrow(cen_col_diff) > 0) {
    cat(sprintf(
      "\n\n**%d value differences** exceed the tolerance (0.001):\n\n",
      nrow(cen_col_diff)
    ))
    cen_col_diff |>
      mutate(across(where(is.numeric), ~ round(.x, 4))) |>
      kable(
        format = "html",
        col.names = c("Geo unit", "Variable", "New", "Legacy",
                       "Diff", "Abs diff", "Within tol"),
        align = "llrrrrr"
      ) |>
      kable_styling(bootstrap_options = c("striped", "condensed"),
                    font_size = 11, full_width = FALSE) |>
      scroll_box(height = "260px") |>
      as.character() |> cat()
  } else {
    cat("\n\n*All values match within tolerance.*\n")
  }
}
```

## Individual-level summary

```{r cen_individual, echo=FALSE}
if (!is.null(cen_ind_sum)) {
  cen_ind_sum |>
    mutate(across(where(is.numeric), ~ round(.x, 2))) |>
    kable(
      format = "html",
      col.names = c("Pipeline", "Total rows", "Adults (25+)",
                     "Weighted N", "Mean schooling", "Share women",
                     "Geo units"),
      align = "lrrrrrr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
}
```

---

# Metropolitan area boundaries {#metro-area}

This section compares the geographic definition of the Bogotá metropolitan area
between the two pipelines and visualises the station networks within each boundary.

## Summary comparison

```{r metro_summary, echo=FALSE, results='asis'}
if (is.null(metro_summary)) {
  cat(paste0(
    status_badge(FALSE, label_pending = "Not yet run"),
    '\n\nRun `compare_metro_area()` in `compare_bogota.R` to generate results.\n\n'
  ))
} else {
  metro_summary |>
    kable(
      format = "html",
      col.names = c("Metric", "Dropbox legacy", "New pipeline"),
      align = "lrr"
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) |>
    column_spec(1, bold = TRUE) |>
    as.character() |> cat()
}
```

## Interactive maps

```{r metro_load, echo=FALSE}
#| include: false
if (metro_has_sf) {
  new_metro_sf       <- sf::st_read(file.path(metro_dir, "new_metro.gpkg"), quiet = TRUE)
  legacy_metro_sf    <- sf::st_read(file.path(metro_dir, "legacy_metro.gpkg"), quiet = TRUE)
  new_stations_sf    <- sf::st_read(file.path(metro_dir, "new_stations.gpkg"), quiet = TRUE)
  legacy_stations_sf <- sf::st_read(file.path(metro_dir, "legacy_stations.gpkg"), quiet = TRUE)

  tile_url  <- "https://cartodb-basemaps-a.global.ssl.fastly.net/light_all/{z}/{x}/{y}.png"
  tile_attr <- paste0(
    '&copy; <a href="https://carto.com/attributions">CARTO</a> | ',
    '&copy; <a href="https://www.openstreetmap.org/copyright">OSM</a>')
}
```

### Dropbox legacy metro area

```{r metro_map_legacy, echo=FALSE, fig.height=5}
if (metro_has_sf) {
  bb <- as.numeric(sf::st_bbox(legacy_metro_sf))
  leaflet::leaflet(height = 450) |>
    leaflet::addTiles(urlTemplate = tile_url, attribution = tile_attr) |>
    leaflet::fitBounds(bb[1], bb[2], bb[3], bb[4]) |>
    leaflet::addPolygons(
      data = legacy_metro_sf,
      color = "#1E3A8A", weight = 1.5,
      fillColor = "#3B82F6", fillOpacity = 0.2
    ) |>
    leaflet::addCircleMarkers(
      data = legacy_stations_sf,
      radius = 5, stroke = TRUE, color = "#111", weight = 1,
      fillColor = "#DC2626", fillOpacity = 0.85,
      label = if ("station_name" %in% names(legacy_stations_sf))
        legacy_stations_sf$station_name else
        if ("station_std" %in% names(legacy_stations_sf))
          legacy_stations_sf$station_std else NULL
    ) |>
    leaflet::addControl(
      html = sprintf(
        '<div style="background:rgba(255,255,255,.85);padding:6px 10px;
         border-radius:6px;font-size:13px">
         <b>Dropbox legacy</b><br>Stations: <b>%d</b></div>',
        nrow(legacy_stations_sf)),
      position = "topleft"
    )
}
```

### New pipeline metro area

```{r metro_map_new, echo=FALSE, fig.height=5}
if (metro_has_sf) {
  bb <- as.numeric(sf::st_bbox(new_metro_sf))
  leaflet::leaflet(height = 450) |>
    leaflet::addTiles(urlTemplate = tile_url, attribution = tile_attr) |>
    leaflet::fitBounds(bb[1], bb[2], bb[3], bb[4]) |>
    leaflet::addPolygons(
      data = new_metro_sf,
      color = "#1E3A8A", weight = 1.5,
      fillColor = "#3B82F6", fillOpacity = 0.2
    ) |>
    leaflet::addCircleMarkers(
      data = new_stations_sf,
      radius = 5, stroke = TRUE, color = "#111", weight = 1,
      fillColor = "#16A34A", fillOpacity = 0.85,
      label = if ("station_name" %in% names(new_stations_sf))
        new_stations_sf$station_name else NULL
    ) |>
    leaflet::addControl(
      html = sprintf(
        '<div style="background:rgba(255,255,255,.85);padding:6px 10px;
         border-radius:6px;font-size:13px">
         <b>New pipeline</b><br>Stations: <b>%d</b></div>',
        nrow(new_stations_sf)),
      position = "topleft"
    )
}
```

---

# Station-to-locality distances {#distances}

This section compares the distance matrices used for inverse-distance weighting
(IDW) between the two pipelines. The new pipeline utilizes an Azimuthal Equidistant
(AEQD) projection for highly accurate planar distances, while the legacy code utilized 
the Haversine great-circle formula. At metro scales (~50–200 km), these coordinate methods
agree within approximately 0.1–0.5 km.

### Spatial Centroid Algorithms and Polygon Concavity

Beyond coordinate reference systems, distance calculations rely on how the "center" of a 
geographic unit is defined. In irregular, concave polygons (such as crescent-shaped census
tracts), a standard mathematical centroid (`st_centroid`) will often fall completely outside
the physical boundaries of the polygon.

To prevent calculating distances from a point outside the locality, both pipelines employ a 
topological fallback:
1.  **Primary:** Calculate the mathematical centroid.
2.  **Validation:** Check if the centroid intersects the polygon boundary.
3.  **Fallback:** If the centroid falls outside, utilize `st_point_on_surface()` to guarantee a
coordinate physically located inside the tract.

Unlike a nearest-boundary projection—which would erroneously anchor the geographic center to 
the extreme edge of the tract (often along boundary highways or rivers)—`st_point_on_surface()` 
utilizes a medial-axis algorithm to find a point deep within the polygon's interior area. This 
guarantees that the representative coordinate is surrounded by the tract's geometry, serving as 
a vastly more accurate proxy for the geographic center of the actual residential population.

**Methodological Note on Legacy Code:** During validation, we identified a vectorization error 
in the legacy script's spatial fallback algorithm. The legacy code evaluated the `st_within()` 
condition strictly on the *first* locality in the dataset and applied that logical outcome 
globally to all subsequent localities (`in_poly <- st_within(...)[[1]]`). Consequently, if the 
first locality was well-shaped, the fallback was disabled for all concave localities; if the 
first was concave, the fallback was applied to perfectly square localities. 

The new automated pipeline corrects this by strictly evaluating the topology row-by-row, 
ensuring maximum spatial fidelity for exposure calculations.

## Station-to-station distances

```{r idw_sta, echo=FALSE, results='asis'}
if (is.null(idw_sta_sum)) {
  cat(paste0(
    status_badge(FALSE, label_pending = "Not yet run"),
    '\n\nRun `compare_idw()` in `compare_bogota.R` to generate results.\n\n'
  ))
} else {
  idw_sta_sum |>
    mutate(
      share_match   = scales::percent(share_match, 0.01),
      mean_abs_diff = round(mean_abs_diff, 3),
      max_abs_diff  = round(max_abs_diff, 3)
    ) |>
    kable(
      format = "html",
      col.names = c("Pairs", "Match", "Diff", "Mean abs diff (km)",
                     "Max abs diff (km)", "Match rate"),
      align = "rrrrrr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) |>
    as.character() |> cat()

  if (!is.null(idw_sta_diff) && nrow(idw_sta_diff) > 0) {
    cat(sprintf(
      "\n\n**%d station pairs** differ by more than the tolerance:\n\n",
      nrow(idw_sta_diff)
    ))
    idw_sta_diff |>
      head(20) |>
      mutate(across(where(is.numeric), ~ round(.x, 3))) |>
      select(station_from, station_to, distance_km_new,
             distance_km_legacy, diff_km) |>
      kable(
        format = "html",
        col.names = c("From", "To", "New (km)", "Legacy (km)", "Diff (km)"),
        align = "llrrr"
      ) |>
      kable_styling(bootstrap_options = c("striped", "condensed"),
                    font_size = 11, full_width = FALSE) |>
      scroll_box(height = "260px") |>
      as.character() |> cat()
  } else {
    cat("\n\n*All station-to-station distances match within tolerance.*\n")
  }
}
```

## Geo-unit to station distances

```{r idw_geo, echo=FALSE, results='asis'}
if (!is.null(idw_geo_sum)) {
  idw_geo_sum |>
    mutate(
      share_match   = scales::percent(share_match, 0.01),
      mean_abs_diff = round(mean_abs_diff, 3),
      max_abs_diff  = round(max_abs_diff, 3)
    ) |>
    kable(
      format = "html",
      col.names = c("Pairs", "Match", "Mean abs diff (km)",
                     "Max abs diff (km)", "Match rate"),
      align = "rrrrr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) |>
    as.character() |> cat()
} else {
  cat("*Geo-unit distance comparison not available or not yet run.*\n")
}
```

> **Methodological note.** The new pipeline computes all distances using AEQD
> (Azimuthal Equidistant) projection centred on the station cloud centroid,
> yielding accurate Euclidean distances. The legacy uses `geosphere::distm()`
> with the Haversine formula. Additionally, the new pipeline identifies
> pollutant-specific nearest neighbours (only stations with data for the
> current pollutant are candidates), while the legacy uses a single nearest
> neighbour for all pollutants.

---

# Outlier-corrected pollution panel {#processed-panel}

Both pipelines apply the same outlier detection algorithm. The procedure works
in three steps for each pollutant, within each station × year-month:

1. **Flag** observations above the 99th percentile.
2. **Benchmark** each flagged observation against its lag/lead average. Classify
   as reasonable (within mean ± 2 SD of first differences) or unreasonable. If
   neither lag nor lead exist, the observation has no benchmark.
3. **Neighbour check** — unreasonable observations are reclassified as reasonable
   if the difference with the nearest neighbour falls within mean ± 2 SD of the
   neighbour-difference distribution.

Flagged observations that remain unreasonable or lack a benchmark are set to NA.

## Outlier detection summary (new pipeline)

```{r outlier_step, echo=FALSE, results='asis'}
if (is.null(out_step_sum)) {
  cat(paste0(
    status_badge(FALSE, label_pending = "Not yet run"),
    '\n\nRun `compare_outlier_procedure()` in `compare_bogota.R` to generate results.\n\n'
  ))
} else {
  out_step_sum |>
    mutate(pct_removed = paste0(pct_removed, "%")) |>
    kable(
      format = "html",
      col.names = c("Pollutant", "Total non-NA obs",
                    "Outliers removed", "% removed",
                    "Missing Temporal (Type 3)", 
                    "Failed Temp + Miss Spat", 
                    "Failed Temp + Failed Spat"),
      align = "lrrrrrr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) |>
    as.character() |> cat()
}
```

## Per-station outlier comparison

```{r outlier_compare, echo=FALSE, results='asis'}
if (!is.null(out_compare) && nrow(out_compare) > 0) {
  cat(paste0(
    "Per-station comparison of outlier counts between the new pipeline and ",
    "the Dropbox legacy:\n\n"
  ))

  out_compare |>
    select(any_of(c("station", "pollutant",
                     "outliers_new", "outliers_legacy", "total_new"))) |>
    arrange(pollutant, station) |>
    kable(
      format = "html",
      col.names = c("Station", "Pollutant",
                     "Outliers (new)", "Outliers (legacy)", "Total obs (new)"),
      align = "llrrr",
      format.args = list(big.mark = ",")
    ) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 11, full_width = FALSE) |>
    scroll_box(height = "300px") |>
    as.character() |> cat()
} else {
  cat("*Per-station comparison not available (legacy cleaned data not found).*\n")
}
```

> **Methodological note.** The new pipeline identifies the nearest neighbour
> on a per-pollutant basis (only stations with data for the current pollutant
> are candidates), while the legacy uses a single nearest neighbour for all
> pollutants. The new pipeline also uses AEQD distances vs the legacy's
> Haversine. These differences may cause a small number of observations to be
> classified differently at step 3.

---

<details><summary>Run metadata</summary>
```{r run_meta}
#| echo: false
tibble::tibble(
  Item = c(
    "City", "Year(s)", "Pipeline timezone",
    "Dropbox legacy source (used in this report)",
    "Dropbox merged CSV (not used — contains Stata bug)",
    "New pipeline source",
    "New pipeline created by",
    "Focus pollutants",
    "New pipeline PM10 stations",
    "New pipeline PM25 stations",
    "Dropbox legacy RMCAB stations"
  ),
  Value = c(
    params$city_id,
    yr_label,
    cmp$pipeline_tz %||% cfg$tz,
    "data/pollution/Bogota/original_data/2020-2023/ (individual period CSVs)",
    "data/Bogota/working_data/Air_Pollution_Bogota_2002_2023.csv",
    "data/raw/monitoring_stations/bogota_metro_dataset/",
    paste0(
      "scripts/download_data/download_bogota_data.R + ",
      "scripts/process_data/process_bogota_data.R ",
      "(bogota_process_stations_data_to_parquet())"
    ),
    paste(toupper(focus_pol), collapse = ", "),
    "48", "47", "19"
  )
) |>
  kable(col.names = c("Item", "Value")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
</details>
