---
title: "Bogotá — Pipeline Validation Report"
subtitle: "Comparison: New Automated Pipeline vs. Coauthor Legacy Dataset"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    self-contained: true
    code-fold: true
    fig-width: 9
    fig-height: 4
params:
  city_id:  "bogota"
  out_root: "results/validation_rep_package"
---

```{r setup, include=FALSE}
library(here); library(arrow); library(dplyr)
library(ggplot2); library(knitr); library(kableExtra)
library(scales); library(lubridate)

source(here("src", "general_utilities",
            "config_utils_validation_old_version.R"))
source(here("src", "city_specific", "registry.R"))
source(here("src", "city_specific", "bogota.R"))

`%||%` <- function(a, b) if (!is.null(a)) a else b

cfg      <- city_cfg(params$city_id)
cmp      <- cfg$compare
city_dir <- file.path(params$out_root, params$city_id)

# Helper: read a Parquet from a named subdirectory
read_pq <- function(subdir, name) {
  f <- file.path(city_dir, subdir, paste0(name, ".parquet"))
  if (file.exists(f)) arrow::read_parquet(f) else NULL
}

# Ground station artefacts
gs_diff_sum  <- read_pq("ground_station_comparison", "diff_summary")
gs_diffs     <- read_pq("ground_station_comparison", "diffs_long")
gs_only_leg  <- read_pq("ground_station_comparison", "only_legacy")
gs_only_new  <- read_pq("ground_station_comparison", "only_new")
gs_sta_audit <- read_pq("ground_station_comparison", "station_audit")

focus_pol <- cmp$focus_pollutants %||% cmp$value_cols
yr_label  <- paste(cmp$compare_years, collapse = ", ")

pol_col <- c(pm10 = "#2b6ca8", pm25 = "#d94f3d",
             ozone = "#e6a817", co = "#4caf50", no2 = "#9c27b0")

# Status badge helper
status_badge <- function(ok, label_ok = "Complete",
                          label_pending = "Pending") {
  if (ok) {
    sprintf(
      '<span style="background:#2e7d32;color:white;padding:2px 8px;
       border-radius:4px;font-size:.85em">%s</span>', label_ok
    )
  } else {
    sprintf(
      '<span style="background:#9e9e9e;color:white;padding:2px 8px;
       border-radius:4px;font-size:.85em">%s</span>', label_pending
    )
  }
}
```

# City overview

> **Bogotá D.C.** is the capital of Colombia and one of the largest cities
> in Latin America. Air quality monitoring is operated by RMCAB
> (Red de Monitoreo de Calidad del Aire de Bogotá), supplemented by
> SISAIRE (the national network) for the broader metropolitan area.
> The comparison covers **`r yr_label`** using PM10 and PM25 as the
> focus pollutants of the project.

---

## Comparison checklist

```{r checklist, echo=FALSE, results='asis'}
checks <- list(
  list(
    name    = "Ground station hourly data",
    done    = !is.null(gs_diff_sum),
    section = "#ground-stations"
  ),
  list(
    name    = "Census microdata",
    done    = FALSE,
    section = "#census"
  ),
  list(
    name    = "Metropolitan area boundaries",
    done    = FALSE,
    section = "#metro-area"
  ),
  list(
    name    = "Station-to-locality distances",
    done    = FALSE,
    section = "#distances"
  ),
  list(
    name    = "Outlier-corrected pollution panel",
    done    = FALSE,
    section = "#processed-panel"
  )
)

rows <- sapply(checks, function(x)
  sprintf("| [%s](%s) | %s |",
          x$name, x$section, status_badge(x$done))
)
cat("| Data layer | Status |\n|---|---|\n")
cat(paste(rows, collapse = "\n"))
```

---

# Ground stations {#ground-stations}

```{r gs_verdict, echo=FALSE, results='asis'}
if (is.null(gs_diff_sum)) {
  cat(
    '<div style="background:#f5f5f5;padding:12px;border-radius:4px">',
    status_badge(FALSE, label_pending = "Not yet run"),
    ' Run <code>compare_bogota.R</code> to generate results.</div>'
  )
  knitr::knit_exit()
}

pm_sum      <- gs_diff_sum |> filter(variable %in% focus_pol)
n_cell_diff <- sum(pm_sum$n_diff)
n_total     <- sum(pm_sum$n)
pct_agree   <- 1 - n_cell_diff / n_total

vcol  <- if (pct_agree >= 0.999) "#2e7d32" else
          if (pct_agree >= 0.990) "#f57c00" else "#c62828"
vtext <- if (pct_agree >= 0.999) "Essentially identical" else
          if (pct_agree >= 0.990) "Minor discrepancies" else
          "Significant differences"

cat(sprintf(paste0(
  '<div style="border-left:5px solid %s;padding:12px 18px;',
  'background:#f9f9f9;border-radius:4px;margin-bottom:16px">',
  '<span style="font-size:1.1em;font-weight:bold;color:%s">%s</span>',
  '<br><span style="color:#555">',
  'PM10 + PM25: <strong>%.4f%%</strong> of matched rows agree exactly. ',
  '<strong>%s</strong> cell differences | ',
  '<strong>%s</strong> rows only in new | ',
  '<strong>%s</strong> rows only in legacy.</span></div>'),
  vcol, vcol, vtext,
  100 * pct_agree,
  format(n_cell_diff,      big.mark = ","),
  format(nrow(gs_only_new), big.mark = ","),
  format(nrow(gs_only_leg), big.mark = ",")))
```

## Pollutant summary

```{r gs_summary_tbl, echo=FALSE}
gs_diff_sum |>
  filter(variable %in% focus_pol) |>
  mutate(
    share_diff    = percent(share_diff, 0.0001),
    mean_abs_diff = round(mean_abs_diff, 3),
    p95_abs_diff  = round(p95_abs_diff,  3)
  ) |>
  kable(
    col.names = c("Pollutant", "Matched rows", "Rows differ",
                  "Both NA", "Share differ",
                  "Mean |diff| (μg/m³)", "P95 |diff| (μg/m³)"),
    align = "lrrrrrr"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  column_spec(5, bold = TRUE)
```

## Cell-level differences

```{r gs_cells, echo=FALSE, results='asis', warning=FALSE}
cell_diffs <- gs_diffs |>
  filter(!within_tol, !is.na(absv), absv > 0, variable %in% focus_pol)

if (nrow(cell_diffs) == 0) {
  cat("*No cell-level differences for PM10 and PM25.*\n")
} else {
  ut <- cell_diffs |> distinct(month, day, hour)
  cat(sprintf(paste0(
    "**%s differing rows** across **%s stations**. ",
    "All fall on month %d, day %d, hour %02d:00 — a single timestamp ",
    "across all stations simultaneously, consistent with an RMCAB ",
    "**data revision** between download dates (mean |diff| = %.1f μg/m³).\n\n"),
    format(nrow(cell_diffs), big.mark = ","),
    dplyr::n_distinct(cell_diffs$station),
    ut$month[1], ut$day[1], ut$hour[1],
    mean(abs(cell_diffs$diff))
  ))
  cell_diffs |>
    select(variable, station, month, day, hour,
           value_old, value_new, diff) |>
    mutate(across(c(value_old, value_new, diff), ~ round(.x, 2))) |>
    arrange(variable, station) |>
    kable(col.names = c("Pollutant", "Station", "Month",
                        "Day", "Hour", "Legacy", "New", "Diff")) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  font_size = 12, full_width = FALSE) |>
    scroll_box(height = "280px") |>
    print()
}
```

## Rows only in the new pipeline

```{r gs_only_new, echo=FALSE, results='asis'}
if (!is.null(gs_only_new) && nrow(gs_only_new) > 0) {
  gs_only_new |>
    mutate(cause = case_when(
      month == 12 & day >= 29     ~ "Dec 31 — 363-day RMCAB cap",
      month == 1  & day == 1 & hour == 0
                                  ~ "Jan 1 00:00 — prior-year data absent",
      month == 2  & day == 1 & hour == 0
                                  ~ "Feb 1 00:00 — prior-year data absent",
      station == "MOVILFONTIBON"  ~ "MOVILFONTIBON ozone-only (no PM)",
      TRUE                        ~ "Other"
    )) |>
    count(cause, name = "rows") |>
    arrange(desc(rows)) |>
    kable(caption = sprintf(
            "%s rows only in new pipeline",
            format(nrow(gs_only_new), big.mark = ",")),
          col.names = c("Cause", "Rows")) |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    print()

  cat(paste0(
    '\n\n> **Root causes.** All three causes stem from differences in ',
    '*download coverage*, not data processing. (1) The RMCAB portal caps ',
    'manual exports at 363 days; our scraper downloads in two half-year ',
    'batches. (2) Stations without prior-year data cannot produce the ',
    'midnight rollover (hour 24 → day+1 hour 0) for Jan 1 and Feb 1. ',
    '(3) MOVILFONTIBON ozone rows are irrelevant for PM analysis.\n\n'
  ))
}
```

## Station coverage

```{r gs_stations, echo=FALSE}
if (!is.null(gs_sta_audit)) {
  gs_sta_audit |>
    mutate(Status = case_when(
      in_legacy & in_new & !sisaire_only ~ "\u2713 Matched",
      sisaire_only                        ~ "\u26a0 SISAIRE-only (filtered)",
      in_legacy & !in_new               ~ "\u2717 Legacy only",
      TRUE                               ~ "\u2295 New only"
    )) |>
    select(station, Status) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE) |>
    row_spec(which(gs_sta_audit$sisaire_only), background = "#fff3cd")
}
```

---

# Census {#census}

```{r census_placeholder, echo=FALSE, results='asis'}
cat(paste0(
  status_badge(FALSE, label_pending = "Pending"),
  '\n\n*Census comparison not yet implemented. ',
  'Add `compare_census()` to `compare_bogota.R` ',
  'and re-render this report.*\n\n'
))
```

---

# Metropolitan area boundaries {#metro-area}

```{r metro_placeholder, echo=FALSE, results='asis'}
cat(paste0(
  status_badge(FALSE, label_pending = "Pending"),
  '\n\n*Metro area boundary comparison not yet implemented.*\n\n'
))
```

---

# Station-to-locality distances {#distances}

```{r dist_placeholder, echo=FALSE, results='asis'}
cat(paste0(
  status_badge(FALSE, label_pending = "Pending"),
  '\n\n*Distance matrix comparison not yet implemented.*\n\n'
))
```

---

# Outlier-corrected pollution panel {#processed-panel}

```{r panel_placeholder, echo=FALSE, results='asis'}
cat(paste0(
  status_badge(FALSE, label_pending = "Pending"),
  '\n\n*Processed panel comparison not yet implemented.*\n\n'
))
```

---

<details><summary>Run metadata</summary>
```{r run_meta, echo=FALSE}
tibble::tibble(
  Item  = c("City", "Year(s)", "Pipeline tz",
            "Legacy source", "Focus pollutants"),
  Value = c(
    params$city_id,
    yr_label,
    cmp$pipeline_tz %||% cfg$tz,
    if (!is.null(cmp$legacy_single_csv) &&
        file.exists(cmp$legacy_single_csv))
      basename(cmp$legacy_single_csv) else cmp$legacy_dir,
    paste(focus_pol, collapse = ", ")
  )
) |>
  kable() |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
</details>
